% \begin{center}
%     \textit{Festina lente.}
% \end{center}
% \begin{flushright}
%     ----- Augustus
% \end{flushright}

%% PI, committee member, lab
This endeavor would not have been possible without my thesis advisor, Dr. Adam Siepel, who has generously provided crucial guidance and feedback throughout my thesis project. I would also like to express my deepest gratitude to my thesis committee who provided their extensive knowledge and expertise. I would like to thank Dr. Jesse Gillis for spending the extra effort to keep my progress on track as the chair of my committee, Dr. David McCandlish for devoting long hours to explaining population genetics theory and giving me career advice as my academic mentor, Dr. Peter Koo for his comprehensive guidance on the world of deep learning, and Dr. Andy Kern for taking interest in my work and being an advocate at many conferences. My collaborators have also been indispensable to the success of my research projects. I would like to express my sincere gratitude to Drs. Leonardo Campagna, Nandita Garud, Hussein Hejase, Rob Martienssen, Patrick Reilly, Armin Scheben, Serena Tucci, Al Uy, Jeremiah Wander, as well as Mariana Harris and Josh Steinberg, In addition, I would like to thank the generous support from the Gladys \& Roland Harriman Fellowship for funding my PhD at Cold Spring Harbor Laboratory.

% school
I am genuinely grateful for the consistent support from everyone at the School of Biological Sciences. Thanks to Kim Creteur, Alex Gann, Kim Graham, Alyson Kass-Eisler, Zach Lippman, Monn Monn Myat, Victoria Panebianco and Catherine Perez for making a difficult PhD as smooth as possible.

I would like to extend my sincere gratitude to all current and past Siepel lab members. In particular, I am indebted to Hussein Hejase for his mentorship at the outset of my PhD, and grateful for the companionship from Katie Brenner, Noah Dukler, Ling Liu, Luiz Machado, Mehreen Mughal, Ritika Ramani, Armin Scheben and Xander Xue. Special thanks to Susan Fredericks for keeping the spirits of the office high, and Melissa Hubisz for her kind help on ARGweaver. I am also thankful for members of my cohort, Alexa, Amritha, Asad, Connor, Dani, Ilgin, Jenelys, Jonathan, Marie and Teri for going through some of the most difficult parts of this journey together. I'd like to give a special shout-out to folks at the Thursday board game club, especially Cole and Michael, for providing a fun and thoughtful break from a busy work day.

Last, but by no means least, words cannot express my gratitude to my family for their unwavering support, both spiritually and financially. To my mom Dq, my dad Hui, aunt Daisy, aunt Jianying, grandpa John, grandma Lily, grandpa Aimin and grandma Guizhen, thank you for your unconditional love.
